---
title: "Assignment 3: Data Exploration"
author: "Reino Hyyppa, Section #02"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
# check wd
getwd()

# load packages
library(tidyverse)

# initialize data
Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)

# initialize data
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: From my understanding, harmful chemicals used in neonicotinoids are dangerous to insects, including pollinators which are important in sustaining a healthy ecosystem. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Litter and woody debris that falls to the forest floor can be beneficial to health of the forest and the wildlife that live in them. In addition, wood debris is a fuel source for widlfires, and monitoring litter and woody debris can help with wildfire management. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* Litter and fine woody debris are only sampled in tower plots.
* Ground traps are sampled once per year.
* Trap placement within plots can be either targeted or randomized, depending on the vegetation. 


## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
# check dimension of Neonics dataset
dim(Neonics)
```
> Answer: Rows: 4623; Columns: 30

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect)
```

> Answer: The effects of neonics are important to study in order to determine the predomoninant effects that neonics have on insects health and survival.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: The six most commonly studies insects are Honey bee, parasitic wasp, buff tailed bumblebee, Carniolan Honey Bee,  Bumble Bee, Italian Honeybee.All six of these species are pollinators. One reason why pollinators might be of greater interest than other insects, is the role that pollinators play in overall ecosystem health. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
Neonics$Conc.1..Author.[1:6]
class(Neonics$Conc.1..Author.)

```

> Answer: Conc.1..Author. is classified as a factor. This variable is stored as a factor because it seems like there's a "/" stored in a lot of the concentration values. Because of this, R is intepretating concentrations as a factor instead of a numeric value. 

## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics, aes(x = Publication.Year))+geom_freqpoly()
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics, aes(x = Publication.Year, color = Test.Location))+geom_freqpoly()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
endpoints <- table(Neonics$Endpoint)

barplot(endpoints, main = "Endpoint Counts",
        xlab = "Endpoint")
```

> Answer: Based on the barchart plot, the two most common endpoints are NOEl and LOEL. They are defined as no-observable-effect level and lowest-observable-effect-level, respectively. 
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
# check class
class(Litter$collectDate)
# set date
Litter$collectDate <- as.Date(Litter$collectDate) 

# use unique function to determine 
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
df_plotID <- unique(Litter$plotID)
length(df_plotID)

summary(df_plotID)
```

> Answer: The unique function produces individual results for each plot samples at Niwot Ridge, whereas, the summary function summarizes all of the plots into a single count. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

LitterType <- table(Litter$functionalGroup)

barplot(LitterType, main = "Litter Types",
        xlab = "Litter")

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

x1 <- Litter$dryMass
x2 <- Litter$functionalGroup

# create a boxplot
ggplot(Litter, aes(x=dryMass, y=functionalGroup)) + 
  geom_boxplot()

# create a violin plot 
ggplot(Litter, aes(x=dryMass, y=functionalGroup)) + 
  geom_violin()
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: A boxplot is a more effective visualization tool for several reasons. First, off a boxplot shows the full distribution of the data and, therefore, outliers are clearly identifiable. Additionally, a boxplot displays summary statistics including median and the interquartile range which can be useful to analyze your data.  

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles and twigs/branches tend to have the highest biomas as these sights.